---
title: "Ebola Forecasting - Residual Analysis"
author: "Andy Shen"
date: "9/11/2020"
output: 
  pdf_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include = FALSE}
## Complete Outbreak Dataset 
## Used to plot the actual number of infections over time
rm(list=ls())
actual <- read.csv("/Volumes/GoogleDrive/.shortcut-targets-by-id/15UGkfREtfqH3LdfHmCsSpFJ5SrTnSeyt/ebola/2020-05-04_ebola/2020-05-04_data.csv")
source("outbreak_vis.R") #script with functions
rgx <- "\\d{1,2}\\/\\d{1,2}\\/\\d{4}" #date structure regex
actual <- actual[str_detect(actual$Date, rgx),] #omits rows without a date
colnames(actual) <- c("date", "cases")
actual$date <- mdy(actual$date) #converts into a consistent date format
actual$cases[is.na(actual$cases)] <- 0
actual <- actual %>% mutate(total = cumsum(cases))
last_date <- actual$date[length(actual$date)]
last_case <- actual$total[length(actual$total)]
```

```{r, include = FALSE}
## Recursive Projections Dataset
rproj <- read.csv("/Volumes/GoogleDrive/.shortcut-targets-by-id/1LaD1nL_OAOposW2fr2XDcs6BLHVBC-jA/2019 Ebola/Post-Outbreak Analysis/recursive_post_proj.csv") #forecasted values
rproj <- rproj %>% select(date_last_case, pred.7, pred.14, pred.21) 
rproj$date_last_case <- mdy(rproj$date_last_case)
rproj <- na.omit(rproj) # some dates have no projections
#rproj <- rproj %>% distinct()
rownames(rproj) <- 1:nrow(rproj)
rdates <- as.character(rproj$date_last_case)
rpreds <- t(rproj %>% select(pred.7,pred.14,pred.21))
omit_h <- c(8,17,22,33,47:49,51,52,55,56,60,64,69,77) #omit from Hawkes because Recursive has no data for these dates
```

```{r, include = FALSE}
## Hawkes Projections Dataset
hproj <- read.csv("/Volumes/GoogleDrive/.shortcut-targets-by-id/1LaD1nL_OAOposW2fr2XDcs6BLHVBC-jA/2019 Ebola/Post-Outbreak Analysis/hawkes_post_proj.csv") #forecasted values
hproj <- hproj %>% select(date_last_case, pred.7, pred.14, pred.21) #selecting only the cols with forecasts
hproj$date_last_case <- mdy(hproj$date_last_case)
hproj <- hproj[-omit_h,] #omit the rows that aren't also in recursive
#hproj <- hproj %>% distinct()
hdates <- as.character(hproj$date_last_case) #for working with dates
hpreds <- t(hproj %>% select(pred.7,pred.14,pred.21)) #take t() for the function
rownames(hproj) <- 1:nrow(hproj)
```

```{r}
dim <- 7
```


# Setup {#setup}

We define the beginning, middle, and end of the outbreak such that the beginning is from May 3, 2018 to September 4, 2018, the middle from September 5, 2018 to August 25, 2019, and the end from August 26, 2019 to May 4, 2020.

# New Ebola Cases by Day

The figure below shows the number of new Ebola cases in West Africa by day during the outbreak. The grey lines represent boundaries between the beginning, middle and end of the outbreak (see section [1](#setup)).

```{r,warning=FALSE}
actual <- actual %>% mutate(seven_day_avg = zoo::rollmean(cases, k = 7,fill = NA))
ggplot(
  data = actual,
  mapping = aes(x = date, y = cases)
) + geom_col(color = "lightblue") + theme_classic() + 
  geom_path(
    data = actual, 
    mapping = aes(x = date, y = seven_day_avg),
    color = "black",
    size = 0.75
  ) +
  theme(plot.caption = element_text(hjust = 0.5)) +
  scale_x_date(date_breaks = "5 months", date_labels = "%b-%y") + 
  geom_vline(aes(xintercept = as.Date("2018-09-04")), col = "grey46") +
  geom_vline(aes(xintercept = as.Date("2019-08-25")), col = "grey46") + 
  theme(
    axis.title.x = element_blank()
  )
ggsave(
  "daily_new_cases.jpg",
  path = "/Users/andyshen/Desktop/Ebola Modeling",
  width = 5.5, height = 5.5, units = "in"
)
```

\pagebreak

# Hawkes vs Recursive 7-Day Residual Plot

```{r,warning=FALSE}
cap <- paste0("Residual plot of Hawkes and Recursive 7-day models")
h7 <- (single_forecast(hdates, hpreds, days = 7)$results)
r7 <- (single_forecast(rdates, rpreds, days = 7)$results)
df <- data.frame(
  date = h7$forecast.date,
  hawkes = h7$resids,
  recursive = r7$resids
)
df <- pivot_longer(df, cols = 2:3, names_to = "forecast", values_to = "resid")
p <- ggplot(
  data = actual,
  mapping = aes(x = date, y = total)
) + 
  theme_classic() + 
  theme(plot.caption = element_text(hjust = 0.5), legend.position = "bottom") +
  scale_x_date(date_breaks = "5 months", date_labels = "%b-%y") + 
  geom_hline(aes(yintercept=0))
  
p + geom_point(
  data = df,
  mapping = aes(x = as.Date(date), y = resid, color = forecast)
) + 
  geom_vline(aes(xintercept = as.Date("2018-09-04")), col = "grey46") +
  geom_vline(aes(xintercept = as.Date("2019-08-25")), col = "grey46")
ggsave(
  "hawkes_vs_rec_7.jpg", 
  path = "/Users/andyshen/Desktop/Ebola Modeling",
  width = dim, height = dim, units = "in"
)
```

\pagebreak




# Hawkes vs Recursive 14-Day Residual Plot

```{r,warning=FALSE}
cap <- paste0("Residual plot of Hawkes and Recursive 14-day models")
h14 <- single_forecast(hdates, hpreds, days = 14)$results
r14 <- single_forecast(rdates, rpreds, days = 14)$results
df <- data.frame(
  date = h14$forecast.date,
  hawkes = h14$resids,
  recursive = r14$resids
)
df <- pivot_longer(df, cols = 2:3, names_to = "forecast", values_to = "resid")
p <- ggplot(
  data = actual,
  mapping = aes(x = date, y = total)
) + 
  theme_classic() + 
  theme(plot.caption = element_text(hjust = 0.5), legend.position = "bottom") +
  scale_x_date(date_breaks = "5 months", date_labels = "%b-%y") + 
  geom_hline(aes(yintercept=0))
  
p + labs(y = "cases") + geom_point(
  data = df,
  mapping = aes(x = as.Date(date), y = resid, color = forecast)
) + 
  geom_vline(aes(xintercept = as.Date("2018-09-04")), col = "grey46") +
  geom_vline(aes(xintercept = as.Date("2019-08-25")), col = "grey46")
ggsave(
  "hawkes_vs_rec_14.jpg", 
  path = "/Users/andyshen/Desktop/Ebola Modeling",
  width = dim, height = dim, units = "in"
)
```

\pagebreak

# Hawkes vs Recursive 21-Day Residual Plot

```{r,warning=FALSE}
h21 <- single_forecast(hdates, hpreds, days = 21)$results
r21 <- single_forecast(rdates, rpreds, days = 21)$results
df <- data.frame(
  date = h21$forecast.date,
  hawkes = h21$resids,
  recursive = r21$resids
)
df <- pivot_longer(df, cols = 2:3, names_to = "forecast", values_to = "resid")
p <- ggplot(
  data = actual,
  mapping = aes(x = date, y = total)
) + 
  theme_classic() + 
  theme(plot.caption = element_text(hjust = 0.5), legend.position = "bottom") +
  scale_x_date(date_breaks = "5 months", date_labels = "%b-%y") + 
  geom_hline(aes(yintercept=0))
  
p + labs(y = "cases") + geom_point(
  data = df,
  mapping = aes(x = as.Date(date), y = resid, color = forecast)
) +
  geom_vline(aes(xintercept = as.Date("2018-09-04")), col = "grey46") +
  geom_vline(aes(xintercept = as.Date("2019-08-25")), col = "grey46")
ggsave(
  "hawkes_vs_rec_21.jpg", 
  path = "/Users/andyshen/Desktop/Ebola Modeling",
  width = dim, height = dim, units = "in"
)
```

\pagebreak




# Hawkes-Only Residual Plot

```{r,warning=F}
p + labs(y = "cases") + 
  geom_point(
    data = h7, 
    mapping = aes(x = as.Date(forecast.date), y = resids, col = "dodgerblue")
  ) +   
  geom_point(
    data = h14, 
    mapping = aes(x = as.Date(forecast.date), y = resids, col = "red2")
  ) + 
  geom_point(
    data = h21, 
    mapping = aes(x = as.Date(forecast.date), y = resids, col = "forestgreen")
  ) + 
  scale_color_identity(
    name = "",
    breaks = c("dodgerblue", "red2", "forestgreen"),
    labels = c("7-day", "14-day", "21-day"),
    guide = "legend"
  ) + geom_vline(aes(xintercept = as.Date("2018-09-04")), col = "grey46") +
  geom_vline(aes(xintercept = as.Date("2019-08-25")), col = "grey46")
ggsave(
  "hawkes_resids.jpg",
  path = "/Users/andyshen/Desktop/Ebola Modeling",
  width = dim, height = dim, units = "in"
)
```

\pagebreak

# Recursive-Only Residual Plot

```{r,warning=F}
p + labs(y = "cases") + 
  geom_point(
    data = r7, 
    mapping = aes(x = as.Date(forecast.date), y = resids, col = "dodgerblue")
  ) +   
  geom_point(
    data = r14, 
    mapping = aes(x = as.Date(forecast.date), y = resids, col = "red2")
  ) + 
  geom_point(
    data = r21, 
    mapping = aes(x = as.Date(forecast.date), y = resids, col = "forestgreen")
  ) + 
  scale_color_identity(
    name = "",
    breaks = c("dodgerblue", "red2", "forestgreen"),
    labels = c("7-day", "14-day", "21-day"),
    guide = "legend"
  ) + geom_vline(aes(xintercept = as.Date("2018-09-04")), col = "grey46") +
  geom_vline(aes(xintercept = as.Date("2019-08-25")), col = "grey46")
ggsave(
  "rec_resids.jpg",
  path = "/Users/andyshen/Desktop/Ebola Modeling",
  width = dim, height = dim, units = "in"
)
```

\pagebreak

# RMSE Calculation During Outbreak


```{r, actual case count}
beg <- actual[as.Date(actual$date) <= as.Date("2018-09-04"),]
mid <- actual[(as.Date(actual$date) > as.Date("2018-09-04")) & (as.Date(actual$date) <= as.Date("2019-08-25")),]
end <- actual[as.Date(actual$date) >= as.Date("2019-08-26"),]
```

```{r, Hawkes}
hproj_beg <- hproj[as.Date(hproj$date_last_case) <= as.Date("2018-09-04"),]
hproj_mid <- hproj[(as.Date(hproj$date_last_case) > as.Date("2018-09-04")) & (as.Date(hproj$date_last_case) <= as.Date("2019-08-25")),]
hproj_end <- hproj[as.Date(hproj$date_last_case) >= as.Date("2019-08-26"),]
```

```{r, Recursive}
rproj_beg <- rproj[as.Date(rproj$date_last_case) <= as.Date("2018-09-04"),]
rproj_mid <- rproj[(as.Date(rproj$date_last_case) > as.Date("2018-09-04")) & (as.Date(rproj$date_last_case) <= as.Date("2019-08-25")),]
rproj_end <- rproj[as.Date(rproj$date_last_case) >= as.Date("2019-08-26"),]
```


## RMSE Calculation

The table below shows the RMSE values from both the Hawkes and Recursive models during the beginning, middle, and end of the outbreak. The dates for each section are defined in section [1](#setup).


```{r}
cap <- paste0("7-Day Hawkes and Recursive RMSE values during the outbreak")
rmse_df <- data.frame(
  Hawkes = c(
    forecast_rmse(hproj_beg$date_last_case, t(hproj_beg[,2:4]), days = 7),
    forecast_rmse(hproj_mid$date_last_case, t(hproj_mid[,2:4]), days = 7),
    forecast_rmse(hproj_end$date_last_case, t(hproj_end[,2:4]), days = 7),
    forecast_rmse(hproj$date_last_case, t(hproj[,2:4]), days = 7)
  ),
  Recursive = c(
    forecast_rmse(rproj_beg$date_last_case, t(rproj_beg[,2:4]), days = 7),
    forecast_rmse(rproj_mid$date_last_case, t(rproj_mid[,2:4]), days = 7),
    forecast_rmse(rproj_end$date_last_case, t(rproj_end[,2:4]), days = 7),
    forecast_rmse(rproj$date_last_case, t(rproj[,2:4]), days = 7)
  )
)
rownames(rmse_df) <- c("Beginning","Middle","End","Total")
rmse_df %>% kable(digits = 3, caption = cap)
```

```{r}
cap <- paste0("14-Day Hawkes and Recursive RMSE values during the outbreak")
rmse_df <- data.frame(
  Hawkes = c(
    forecast_rmse(hproj_beg$date_last_case, t(hproj_beg[,2:4]), days = 14),
    forecast_rmse(hproj_mid$date_last_case, t(hproj_mid[,2:4]), days = 14),
    forecast_rmse(hproj_end$date_last_case, t(hproj_end[,2:4]), days = 14),
    forecast_rmse(hproj$date_last_case, t(hproj[,2:4]), days = 14)
  ),
  Recursive = c(
    forecast_rmse(rproj_beg$date_last_case, t(rproj_beg[,2:4]), days = 14),
    forecast_rmse(rproj_mid$date_last_case, t(rproj_mid[,2:4]), days = 14),
    forecast_rmse(rproj_end$date_last_case, t(rproj_end[,2:4]), days = 14),
    forecast_rmse(rproj$date_last_case, t(rproj[,2:4]), days = 14)
  )
)
rownames(rmse_df) <- c("Beginning","Middle","End","Total")
rmse_df %>% kable(digits = 3, caption = cap)
```

```{r}
cap <- paste0("21-Day Hawkes and Recursive RMSE values during the outbreak")
rmse_df <- data.frame(
  Hawkes = c(
    forecast_rmse(hproj_beg$date_last_case, t(hproj_beg[,2:4]), days = 21),
    forecast_rmse(hproj_mid$date_last_case, t(hproj_mid[,2:4]), days = 21),
    forecast_rmse(hproj_end$date_last_case, t(hproj_end[,2:4]), days = 21),
    forecast_rmse(hproj$date_last_case, t(hproj[,2:4]), days = 21)
  ),
  Recursive = c(
    forecast_rmse(rproj_beg$date_last_case, t(rproj_beg[,2:4]), days = 21),
    forecast_rmse(rproj_mid$date_last_case, t(rproj_mid[,2:4]), days = 21),
    forecast_rmse(rproj_end$date_last_case, t(rproj_end[,2:4]), days = 21),
    forecast_rmse(rproj$date_last_case, t(rproj[,2:4]), days = 21)
  )
)
rownames(rmse_df) <- c("Beginning","Middle","End","Total")
rmse_df %>% kable(digits = 3, caption = cap)
```

It is clear that both models are most accurate in the beginning and end of the pandemic as opposed to the middle portion.
